---
title: "Point Counts"
editor: visual
---
# Point count schedule by month

This page will show you the schedule of each point count by month for each of the three visits needed per month.
```{r}
#| echo: false
#| include: false
#| message: false

#####    load packages and functions    #####

{
  library(dplyr)
  library(ggplot2)
  library(cowplot)
  library(knitr)
  library(tidyr)

}

funk <- function(x){length(unique(x))}

```

```{r}
#| echo: false
#| include: false
#| message: false

#####    load data    #####
#
# load each raw data file
survey <- read.csv("C:\\Users\\rmeyer\\Documents\\niche_quarto\\WoodpeckerNiche.github.io\\docs\\data\\order_of_point_count_surveys.csv")

```

## April
### Arcadia
```{r}
#| echo: false
#| message: false
loc <- "ARCA"
mon <- "April"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Arcadia", "in", mon)))
```

### Greenwood
```{r}
#| echo: false
#| message: false
loc <- "GREN"
mon <- "April"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Greenwood", "in", mon)))
```

### Livingston
```{r}
#| echo: false
#| message: false
loc <- "LIVI"
mon <- "April"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Livingston Place", "in", mon)))
```

### Tall Timbers
```{r}
#| echo: false
#| message: false
loc <- "TTRS"
mon <- "April"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Tall Timbers", "in", mon)))
```

## May

### Arcadia
```{r}
#| echo: false
#| message: false
loc <- "ARCA"
mon <- "May"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Arcadia", "in", mon)))
```

### Greenwood
```{r}
#| echo: false
#| message: false
loc <- "GREN"
mon <- "May"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Greenwood", "in", mon)))
```

### Livingston
```{r}
#| echo: false
#| message: false
loc <- "LIVI"
mon <- "May"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Livingston Place", "in", mon)))
```

### Tall Timbers
```{r}
#| echo: false
#| message: false
loc <- "TTRS"
mon <- "May"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Tall Timbers", "in", mon)))
```

## June

### Arcadia
```{r}
#| echo: false
#| message: false
loc <- "ARCA"
mon <- "June"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Arcadia", "in", mon)))
```

### Greenwood
```{r}
#| echo: false
#| message: false
loc <- "GREN"
mon <- "June"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Greenwood", "in", mon)))
```

### Livingston
```{r}
#| echo: false
#| message: false
loc <- "LIVI"
mon <- "June"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Livingston Place", "in", mon)))
```

### Tall Timbers
```{r}
#| echo: false
#| message: false
loc <- "TTRS"
mon <- "June"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Tall Timbers", "in", mon)))
```

## July

### Arcadia
```{r}
#| echo: false
#| message: false
loc <- "ARCA"
mon <- "July"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Arcadia", "in", mon)))
```

### Greenwood
```{r}
#| echo: false
#| message: false
loc <- "GREN"
mon <- "July"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Greenwood", "in", mon)))
```

### Livingston
```{r}
#| echo: false
#| message: false
loc <- "LIVI"
mon <- "July"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Livingston Place", "in", mon)))
```

### Tall Timbers
```{r}
#| echo: false
#| message: false
loc <- "TTRS"
mon <- "July"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Tall Timbers", "in", mon)))
```

## August

### Arcadia
```{r}
#| echo: false
#| message: false
loc <- "ARCA"
mon <- "August"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Arcadia", "in", mon)))
```

### Greenwood
```{r}
#| echo: false
#| message: false
loc <- "GREN"
mon <- "August"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Greenwood", "in", mon)))
```

### Livingston
```{r}
#| echo: false
#| message: false
loc <- "LIVI"
mon <- "August"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Livingston Place", "in", mon)))
```

### Tall Timbers
```{r}
#| echo: false
#| message: false
loc <- "TTRS"
mon <- "August"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Tall Timbers", "in", mon)))
```
