---
title: "Point Counts"
editor: visual
---
# Point count results

Avian communities are often sampled with point count surveys, and for good reason. Point counts in this study will provide information on the grater cavity-nesting bird community which is important for understanding competition and community assemblages as they relate to environmental variables. Below are some preliminary results on each of the four sites with simple models using the Distance package in R. These models vary detection probability by species only and provide estimates on species abundance with associated confidence intervals.

```{r}
#| echo: false
#| include: false
#| message: false

#####    load packages and functions    #####

{
  library(dplyr)
  library(ggplot2)
  library(cowplot)
  library(knitr)
  library(tidyr)
  library(Distance)
  library(purrr)
  library(lubridate)
  library(ggplot2)
  library(unmarked)
  library(lubridate)

}

funk <- function(x){length(unique(x))}



```

```{r}
#| echo: false
#| include: false
#| message: false

#####    load data    #####
#
# load each raw data file
survey <- read.csv("C:\\Users\\rmeyer\\Documents\\niche_quarto\\WoodpeckerNiche.github.io\\docs\\data\\order_of_point_count_surveys.csv")
pc <- read.csv("C:\\Users\\rmeyer\\Documents\\niche_quarto\\WoodpeckerNiche.github.io\\docs\\data\\raw\\point_count_data.csv")
sites <- read.csv("C:\\Users\\rmeyer\\Documents\\niche_quarto\\WoodpeckerNiche.github.io\\docs\\data\\raw\\point_count_id_site_crosswalk.csv")
truncation_dist <- 200
#edit data to have desired columns
pc <- pc[pc$real_distance_m <= truncation_dist, ]
pc <- pc %>%
  filter(!is.na(real_distance_m), real_distance_m <= truncation_dist) %>%
  filter(!species %in% c("YBSA", "WODU"))

# add new columns
pc <- pc %>%
  # Convert date to Date format
  mutate(date = mdy(date)) %>%
  
  # Create visit number per point_id based on date order
  group_by(point_id, date) %>%
  mutate(visit_date = first(date)) %>%   # Ensures a single date per group
  ungroup() %>%
  arrange(point_id, visit_date) %>%
  group_by(point_id) %>%
  mutate(visit = dense_rank(visit_date)) %>%
  
  # Calculate max number of visits per point_id
  mutate(repeats = (n_distinct(visit_date))) %>% #effort in hours

  mutate(Effort = ((n_distinct(visit_date))*10)/60) %>% #effort in hours
  ungroup()
pc <- na.omit(pc)
pc$sky <- as.factor(pc$sky)
pc$wind <- as.factor(pc$wind)
pc <- merge(x=pc,y=sites,by="point_id")
```

```{r}
#| echo: false
#| include: false
#| message: false

# calculate results from point counts
pc_ds <- data.frame(
  object = 1:nrow(pc),
  Region.Label = as.factor(pc$site_name),
  Sample.Label = as.factor(pc$point_id),
  distance = pc$real_distance_m,
  species = as.factor(pc$species),
  sky = pc$sky,
  wind = pc$wind,
  temp=pc$temp,
  Effort=pc$Effort,
  visit=pc$visit,
  repeats=pc$repeats
)
pc_ds$Area <- 12.57 #ha
pc_ds$object <- seq(1,nrow(pc_ds),1)
pc_ds <- na.omit(pc_ds)
pc_ds$sky <- as.factor(pc_ds$sky)
pc_ds$wind <- as.factor(pc_ds$wind)
# Fit half-normal model with detection covariates
truncation_dist <- 200  # Max detection distance in meters
area_ha <- pi * (truncation_dist^2) / 10000  # Survey area in hectares


#model to see what set of variables are best
convunit <- convert_units("meter", NULL, "hectare") #NULL is for point counts
#all.birds.swt <- ds(data = pc_ds, key="hn", convert_units = convunit,transect = "point",
#                formula=~species+sky+wind+temp, truncation = truncation_dist)

#all.birds.sw  <- ds(data = pc_ds, key="hn", convert_units = convunit,transect = "point",
#                formula=~species+sky+wind, truncation = truncation_dist)

#all.birds.st  <- ds(data = pc_ds, key="hn", convert_units = convunit,transect = "point",
#                formula=~species+sky+temp, truncation = truncation_dist)

#all.birds.tw  <- ds(data = pc_ds, key="hn", convert_units = convunit,transect = "point",
#                formula=~species+wind+temp, truncation = truncation_dist)

#all.birds.s   <- ds(data = pc_ds, key="hn", convert_units = convunit,transect = "point",
#                formula=~species+sky, truncation = truncation_dist)

#all.birds.t   <- ds(data = pc_ds, key="hn", convert_units = convunit,transect = "point",
#                formula=~species+temp, truncation = truncation_dist)
#all.birds.w   <- ds(data = pc_ds, key="hn", convert_units = convunit,transect = "point",
#                formula=~species+wind, truncation = truncation_dist)
#all.birds.n   <- ds(data = pc_ds, key="hn", convert_units = convunit,transect = "point",
#                formula=~species, truncation = truncation_dist)

#aic_table <- AIC(all.birds.swt,                             # global model
#                 all.birds.sw,all.birds.st,all.birds.tw,    # 2 cov only models
#                 all.birds.s,all.birds.t,all.birds.w,       # 1 cov only models
#                 all.birds.n                                # null model
#                 )

#aic_table_df <- as.data.frame(aic_table) %>%
#  tibble::rownames_to_column("Model") %>%
#  arrange(AIC)

#print(aic_table_df)

#take best model and subset by site

#best_mod <- ds(data = pc_ds, key="hn", convert_units = convunit,transect = "point",
#                formula=~species, truncation = truncation_dist)



## subset for each site


# Arcadia
pc_ds_a <- subset(pc_ds,pc_ds$Region.Label=="Arcadia")
pc_ds_a$object <- seq(1,nrow(pc_ds_a),1)
arca <- ds(data = pc_ds_a, key="hn", convert_units = convunit,
           transect = "point", truncation = truncation_dist,
           formula=~species)
arca.ests <- dht2(ddf=arca, flatfile=pc_ds,transect = "point",
                  strat_formula = ~species, convert_units = convunit,
                  stratification = "object") 
# Extract abundance estimates as a data frame
temp <- data.frame(
  species = arca.ests$species,
  est = arca.ests$Abundance,
  lower = arca.ests$LCI,
  upper = arca.ests$UCI)

# Separate out the 'Total' row
arca_total <- temp %>% filter(species == "Total")

# Remove the 'Total' row from main data
temp <- temp %>% filter(species != "Total")


# Plot in ggplot2
arca_plot <- ggplot(temp, aes(x = species, y = est)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  labs(title = "Arcadia",
       x = "Species",
       y = "Abundance") +
  scale_y_continuous(limits = c(0, 15), expand = c(0, 0))+
  coord_flip() +  # Optional: flip axes for readability
  theme_bw()

# Greenwood
pc_ds_g <- subset(pc_ds,pc_ds$Region.Label=="Greenwood")
pc_ds_g$object <- seq(1,nrow(pc_ds_g),1)
gren <- ds(data = pc_ds_g, key="hn", convert_units = convunit,
           transect = "point", truncation = truncation_dist,
           formula=~species)

gren.ests <- dht2(ddf=gren, flatfile=pc_ds,transect = "point",
                  strat_formula = ~species, convert_units = convunit,
                  stratification = "object") 
# Extract abundance estimates as a data frame
temp <- data.frame(
  species = gren.ests$species,
  est = gren.ests$Abundance,
  lower = gren.ests$LCI,
  upper = gren.ests$UCI)

# Separate out the 'Total' row
gren_total <- temp %>% filter(species == "Total")

# Remove the 'Total' row from main data
temp <- temp %>% filter(species != "Total")


# Plot in ggplot2
gren_plot <- ggplot(temp, aes(x = species, y = est)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  labs(title = "Greenwood",
       x = "Species",
       y = "Abundance") +
  scale_y_continuous(limits = c(0, 15), expand = c(0, 0))+
  coord_flip() +  # Optional: flip axes for readability
  theme_bw()
# Livingston
pc_ds_l <- subset(pc_ds,pc_ds$Region.Label=="Livingston Place")
pc_ds_l$object <- seq(1,nrow(pc_ds_l),1)
livi <- ds(data = pc_ds_l, key="hn", convert_units = convunit,
           transect = "point", truncation = truncation_dist,
           formula=~species)

livi.ests <- dht2(ddf=livi, flatfile=pc_ds,transect = "point",
                  strat_formula = ~species, convert_units = convunit,
                  stratification = "object") 
# Extract abundance estimates as a data frame
temp <- data.frame(
  species = livi.ests$species,
  est = livi.ests$Abundance,
  lower = livi.ests$LCI,
  upper = livi.ests$UCI)

# Separate out the 'Total' row
livi_total <- temp %>% filter(species == "Total")

# Remove the 'Total' row from main data
temp <- temp %>% filter(species != "Total")


# Plot in ggplot2
livi_plot <- ggplot(temp, aes(x = species, y = est)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  labs(title = "Livingston Place",
       x = "Species",
       y = "Abundance") +
  scale_y_continuous(limits = c(0, 15), expand = c(0, 0))+
  coord_flip() +  # Optional: flip axes for readability
  theme_bw()

# Tall Timbers
pc_ds_t <- subset(pc_ds,pc_ds$Region.Label=="Tall Timbers")
pc_ds_t$object <- seq(1,nrow(pc_ds_t),1)
ttrs <- ds(data = pc_ds_t, key="hn", convert_units = convunit,
           transect = "point", truncation = truncation_dist,
           formula=~species)

ttrs.ests <- dht2(ddf=ttrs, flatfile=pc_ds,transect = "point",
                  strat_formula = ~species, convert_units = convunit,
                  stratification = "object") 
# Extract abundance estimates as a data frame
temp <- data.frame(
  species = ttrs.ests$species,
  est = ttrs.ests$Abundance,
  lower = ttrs.ests$LCI,
  upper = ttrs.ests$UCI)

# Separate out the 'Total' row
ttrs_total <- temp %>% filter(species == "Total")

# Remove the 'Total' row from main data
temp <- temp %>% filter(species != "Total")


# Plot in ggplot2
ttrs_plot <- ggplot(temp, aes(x = species, y = est)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  labs(title = "Tall Timbers",
       x = "Species",
       y = "Abundance") +
  scale_y_continuous(limits = c(0, 15), expand = c(0, 0))+
  coord_flip() +  # Optional: flip axes for readability
  theme_bw()


# arrange data to look at grand totals of cavity nesters across sites

temp <- rbind(arca_total,gren_total,livi_total,ttrs_total)
temp$site <- c("Arcadia","Greenwood","Livingston Place","Tall Timbers")

# Plot in ggplot2
total_plot <- ggplot(temp, aes(x = site, y = est)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  labs(title = "",
       x = "",
       y = "Abundance") +
  scale_y_continuous(limits = c(0, 45), expand = c(0, 0))+
  coord_flip() +  # Optional: flip axes for readability
  theme_bw()


```
## Summary plots for results thus far...

```{r}
#| echo: false
#| message: false

#plotting


#simple plot

#plot(all.birds, showpoints=FALSE, main="Detection distances at point counts\nspecies as covariate")
#add.df.covar.line(all.birds, data=data.frame(species="NOFL"), lwd=3, lty=1, col="blue")
#add.df.covar.line(all.birds, data=data.frame(species="DOWO"), lwd=3, lty=1, col="darkgreen")
#add.df.covar.line(all.birds, data=data.frame(species="RBWO"), lwd=3, lty=1, col="brown")
#add.df.covar.line(all.birds, data=data.frame(species="RCWO"), lwd=3, lty=1, col="salmon")
#add.df.covar.line(all.birds, data=data.frame(species="RHWO"), lwd=3, lty=1, col="red")
#add.df.covar.line(all.birds, data=data.frame(species="PIWO"), lwd=3, lty=1, col="black")
#legend("topright", legend=c("NOFL", "DOWO", "RBWO", "RCWO","RHWO","PIWO"),
#       lwd=3, lty=1, col=c("blue", "darkgreen", "brown", "salmon","red","black"))


#cowplot

plot_grid(arca_plot, gren_plot, livi_plot, ttrs_plot,
          nrow = 2, ncol = 2)#  for letters A-D use: labels = "AUTO"
```

This plot (above) shows the abundance estimates for cavity nesting species at each study site individually. This data represents only March and April 2025.

## Totals across sites

```{r}
#| echo: false
#| message: false

#plotting
total_plot


```

The plot (above) shows total cavity nesting bird abundances at each site.

# How long should a point count be?

In this study we are doing 10min point counts. It could be that 10min is too long, and most of the species we detect will be detected in a shorter amount of time. If so, we may be able to do shorter point counts but we don't want to risk lowering our detection of birds that are more difficult to detect (those that are quiet or rare). Let's look at a simple species accumulation curve...

```{r}
#| echo: false
#| include: false
#| message: false

#how long is long enough for a point count? an analysis in unmarked


# convert mm:ss to numeric minutes
pc_clean <- pc %>%
  mutate(
    time_mins = as.numeric(ms(time_detected)) / 60  
  )

# Step 1: Keep only the FIRST detection per species per visit (point_id + visit)
pc_unique <- pc_clean %>%
  arrange(point_id, visit, species, time_mins) %>%
  group_by(point_id, visit, species) %>%
  slice_head(n = 1) %>%  # keep only first detection per species per visit
  ungroup()

# Step 2: Bin by time again
pc_binned <- pc_unique %>%
  mutate(time_bin = cut(time_mins,
                        breaks = seq(0, 10, by = 2),
                        labels = c("0-2", "2-4", "4-6", "6-8", "8-10"),
                        right = FALSE))

# Step 3: Count number of species per visit per time bin
richness_per_visit <- pc_binned %>%
  group_by(point_id, visit, time_bin) %>%
  summarise(richness = n_distinct(species), .groups = "drop")

# Step 4: Pivot and calculate cumulative richness
richness_wide <- richness_per_visit %>%
  pivot_wider(names_from = time_bin, values_from = richness, values_fill = 0)


#
# Time bin order
time_bin_order <- c("0-2", "2-4", "4-6", "6-8", "8-10")

# Extract just the time bin values, do row-wise cumsum, and bind back
richness_cum <- richness_wide %>%
  select(point_id, visit, all_of(time_bin_order)) %>%
  mutate(across(all_of(time_bin_order), as.numeric))  # ensure numeric just in case

# Apply row-wise cumsum to only the bin columns
cumulative_bins <- t(apply(richness_cum[time_bin_order], 1, cumsum))  # returns a matrix

# Bind back to original identifiers
richness_cum_final <- bind_cols(
  richness_wide %>% select(point_id, visit),
  as.data.frame(cumulative_bins)
)

# Set correct column names
colnames(richness_cum_final)[-(1:2)] <- time_bin_order

# Check
head(richness_cum_final)


#



# Step 5: Average across visits per point
richness_mean_by_plot <- richness_cum_final %>%
  group_by(point_id) %>%
  summarise(across(`0-2`:`8-10`, mean), .groups = "drop")

# Reshape to long format for plotting
richness_long <- richness_mean_by_plot %>%
  pivot_longer(cols = `0-2`:`8-10`, names_to = "TimeBin", values_to = "MeanRichness")

# Ensure time bins are ordered properly
richness_long$TimeBin <- factor(richness_long$TimeBin, levels = c("0-2", "2-4", "4-6", "6-8", "8-10"))

richness_overall <- richness_long %>%
  group_by(TimeBin) %>%
  summarise(MeanRichness = mean(MeanRichness))



```


```{r}
#| echo: false
#| message: false

#plot the curve
ggplot(richness_long, aes(x = TimeBin, y = MeanRichness, group = factor(point_id), color = factor(point_id))) +
  geom_line(alpha = 0.3) +
  geom_line(data = richness_overall, aes(x = TimeBin, y = MeanRichness, group = 1), color = "black", size = 1.5) +
  labs(
    title = "",
    x = "Time Bin (min)",
    y = "Mean Cumulative Species Richness"
  ) +
  scale_y_continuous(breaks = c(0:7),limits = c(0, 7), expand = c(0, 0))+
  theme_bw() +
  theme(legend.position = "none")
```

Looks like modest accumulation with no sharp increase and plateau seen across plots with some variation between plots. The variation between plots could be random noise. 

# Point count schedule by month

The following are the scheduled order of point counts to conduct per visit to the site. This ensures that community data obtained isn't biased by lower or higher detections thoughout the day. 


## April
### Arcadia
```{r}
#| echo: false
#| message: false
loc <- "ARCA"
mon <- "April"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Arcadia", "in", mon)))
```

### Greenwood
```{r}
#| echo: false
#| message: false
loc <- "GREN"
mon <- "April"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Greenwood", "in", mon)))
```

### Livingston
```{r}
#| echo: false
#| message: false
loc <- "LIVI"
mon <- "April"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Livingston Place", "in", mon)))
```

### Tall Timbers
```{r}
#| echo: false
#| message: false
loc <- "TTRS"
mon <- "April"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Tall Timbers", "in", mon)))
```

## May

### Arcadia
```{r}
#| echo: false
#| message: false
loc <- "ARCA"
mon <- "May"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Arcadia", "in", mon)))
```

### Greenwood
```{r}
#| echo: false
#| message: false
loc <- "GREN"
mon <- "May"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Greenwood", "in", mon)))
```

### Livingston
```{r}
#| echo: false
#| message: false
loc <- "LIVI"
mon <- "May"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Livingston Place", "in", mon)))
```

### Tall Timbers
```{r}
#| echo: false
#| message: false
loc <- "TTRS"
mon <- "May"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Tall Timbers", "in", mon)))
```

## June

### Arcadia
```{r}
#| echo: false
#| message: false
loc <- "ARCA"
mon <- "June"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Arcadia", "in", mon)))
```

### Greenwood
```{r}
#| echo: false
#| message: false
loc <- "GREN"
mon <- "June"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Greenwood", "in", mon)))
```

### Livingston
```{r}
#| echo: false
#| message: false
loc <- "LIVI"
mon <- "June"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Livingston Place", "in", mon)))
```

### Tall Timbers
```{r}
#| echo: false
#| message: false
loc <- "TTRS"
mon <- "June"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Tall Timbers", "in", mon)))
```

## July

### Arcadia
```{r}
#| echo: false
#| message: false
loc <- "ARCA"
mon <- "July"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Arcadia", "in", mon)))
```

### Greenwood
```{r}
#| echo: false
#| message: false
loc <- "GREN"
mon <- "July"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Greenwood", "in", mon)))
```

### Livingston
```{r}
#| echo: false
#| message: false
loc <- "LIVI"
mon <- "July"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Livingston Place", "in", mon)))
```

### Tall Timbers
```{r}
#| echo: false
#| message: false
loc <- "TTRS"
mon <- "July"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Tall Timbers", "in", mon)))
```

## August

### Arcadia
```{r}
#| echo: false
#| message: false
loc <- "ARCA"
mon <- "August"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Arcadia", "in", mon)))
```

### Greenwood
```{r}
#| echo: false
#| message: false
loc <- "GREN"
mon <- "August"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Greenwood", "in", mon)))
```

### Livingston
```{r}
#| echo: false
#| message: false
loc <- "LIVI"
mon <- "August"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Livingston Place", "in", mon)))
```

### Tall Timbers
```{r}
#| echo: false
#| message: false
loc <- "TTRS"
mon <- "August"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Tall Timbers", "in", mon)))
```
