---
title: "Point Counts"
editor: visual
---

# Point count results

Avian communities are often sampled with point count surveys, and for good reason. Point counts in this study will provide information on the grater cavity-nesting bird community which is important for understanding competition and community assemblages as they relate to environmental variables. Below are some preliminary results on each of the four sites with simple models using the spAbundance package in R. These models vary detection probability by species only and provide estimates on species abundance with associated confidence intervals.

```{r}
#| echo: false
#| include: false
#| message: false

#####    load packages and functions    #####

{
  library(dplyr)
  library(ggplot2)
  library(cowplot)
  library(knitr)
  library(tidyr)
  
  #library(Distance)
  library(spAbundance)
  library(unmarked)

  library(purrr)
  library(lubridate)
  library(ggplot2)
  #library(unmarked)
  library(lubridate)
  library(abind)
  library(suncalc)
  library(stringr)
}

funk <- function(x){length(unique(x))}



```

```{r spAbundance}
#| echo: false
#| include: false
#| message: false

#####    load data    #####
#
# load each raw data file
survey <- read.csv("C:\\Users\\rmeyer\\Documents\\WoodpeckerNiche\\docs\\data\\order_of_point_count_surveys.csv")
pc <- read.csv("C:\\Users\\rmeyer\\Documents\\WoodpeckerNiche\\docs\\data\\raw\\point_count_data.csv")
sites <- read.csv("C:\\Users\\rmeyer\\Documents\\WoodpeckerNiche\\docs\\data\\raw\\point_count_id_site_crosswalk.csv")
truncation_dist <- 200

# Filter and prepare data

pc <- pc%>%
    mutate(date = mdy(date)) %>%
    filter(date < as.Date("2026-07-01"))%>%                # remove this once I get July data
    filter(!(date >= "2025-07-01" & date <= "2025-08-31")) # remove this once I get July data

pc <- pc %>%
  filter(!is.na(real_distance_m)) %>%
  filter(!species %in% c("YBSA", "WODU","BASP")) %>%
  filter(!species %in% c("HAWO")) %>%        # removing species detected less than 5% of the time, includes NOFL
  filter(date > as.Date("2025-03-31"))%>%            # remove march data
  mutate(julian_day = lubridate::yday(date))%>%          #for consistency
  left_join(sites, by = "point_id") %>%
  group_by(point_id, date) %>%
  mutate(visit_date = first(date)) %>%
  ungroup() %>%
  mutate(double_obs = str_replace_all(double_obs, "n", ""))%>% #for consistency
  mutate(double_obs = tolower(double_obs))%>%                  #for consistency
  mutate(DO_sp_matching = tolower(DO_sp_matching))%>%          #for consistency
  select(-angle,-distance_cm,-comments)%>%
  arrange(point_id, visit_date) %>%
  group_by(point_id) %>%
  mutate(visit = dense_rank(visit_date)) %>%
  ungroup()%>%
  mutate(year=format(date,"%Y"))

#remove second visit (the middle visit in April that is extra)

pc <- pc %>%
  filter(visit!=2) %>%
  arrange(point_id, visit_date) %>%  #re-rank the visits
  group_by(point_id,year) %>%
  mutate(visit = dense_rank(visit_date)) %>%
  ungroup()



# Only keep primary observer’s double observer records
do <- pc %>%
  filter(double_obs == "y", observer == "RTM")

pc <- pc %>%
  filter(double_obs == "") %>%
  bind_rows(do)
pc <- pc%>% select(-double_obs,-DO_sp_matching) #don't need it anymore

# add time since dawn
pc <- pc %>%
  mutate(datetime = ymd_hm(paste(pc$date, pc$start), tz = "America/New_York"))
sun <- getSunlightTimes(
  date = unique(pc$date),
  lat = 30.6,
  lon = -84.1,
  keep = "sunrise",
  tz = "America/New_York"
)
pc <- pc %>%
  left_join(sun, by = "date")
pc <- pc %>%
  mutate(time_since_sunrise =
           as.numeric(difftime(datetime, sunrise, units = "mins")))%>%
  mutate(year=format(date,"%Y"))

#remove EM, does this improve estimates?
#pc <- pc %>%
#  filter(observer == "RTM")

# Encode point_id as numeric
pc$point_id <- as.numeric(pc$point_id)
pc$year <- as.factor(pc$year)
#pc$site_name <- as.numeric(as.factor(pc$site_name))



# combine year and point ID, spAbundance can't handle years otherwise

pc$point <- as.factor(pc$point_id)
pc$point_id <- as.numeric(paste0(pc$year,pc$point_id))
# Define species, sites, and number of replicates
species_list <- unique(pc$species)
site_list <- unique(pc$point_id)
n_species <- length(species_list)
n_sites <- length(site_list)
n_replicates <- pc %>%
  group_by(point_id) %>%
  summarize(n_visits = n_distinct(visit)) %>%
  pull(n_visits) %>%
  max()

# Initialize y array
y_array <- array(0, dim = c(n_species, n_sites, n_replicates),
                 dimnames = list(species = species_list, site = site_list, visit = 1:n_replicates))

# Fill y array with counts per species/site/visit
for (s in seq_along(species_list)) {
  species_name <- species_list[s]
  for (p in seq_along(site_list)) {
    site_id <- site_list[p]
    for (v in 1:n_replicates) {
      y_array[s, p, v] <- pc %>%
        filter(species == species_name, point_id == site_id, visit == v) %>%
        nrow()
    }
  }
}

# check for 0 inflation
# Calculate observed vs expected zeros under Poisson
observed_counts <- y_array[species_list, , ]  # For a specific species
n_zeros_obs <- sum(observed_counts == 0, na.rm = TRUE)
n_total <- sum(!is.na(observed_counts))

# Compare to expected under fitted model
lambda <- mean(observed_counts, na.rm = TRUE)
n_zeros_exp <- n_total * dpois(0, lambda)

cat("Observed zeros:", n_zeros_obs, "\n")
cat("Expected zeros (Poisson):", round(n_zeros_exp), "\n")
cat("Ratio:", n_zeros_obs / n_zeros_exp, "\n")
# Ratio > 1.5 suggests zero-inflation

#detection_counts <- apply(y, 1, sum, na.rm = TRUE)
#keep_species <- detection_counts >= 20  # Adjust threshold
#y_filtered <- y[keep_species, , ]

# Create detection covariates for each point_id × visit

det_covs <- pc %>%
  arrange(point_id, visit)%>%
  group_by(point_id, visit) %>%
  summarize(wind = first(wind),
            sky = first(sky),
            observer = first(observer), 
            time_since_sunrise = first(time_since_sunrise),
            julian_day=first(julian_day),
            .groups = "drop") 

det_covs_matrix <- det_covs %>%
  pivot_wider(names_from = visit, values_from = c(wind, sky, observer,
                                                  time_since_sunrise,julian_day)) %>%
  arrange(point_id)

wind_mat <- det_covs_matrix %>%
  select(starts_with("wind_")) %>%
  as.matrix()
wind_mat[is.na(wind_mat)] <- 0

sky_mat <- det_covs_matrix %>%
  select(starts_with("sky_")) %>%
  as.matrix()
sky_mat[is.na(sky_mat)] <- 0

observer_mat <- det_covs_matrix %>%
  select(starts_with("observer_")) %>%
  as.data.frame()
# as.matrix()
#observer_mat <- apply(observer_mat, 2, function(x) as.numeric(as.factor(x)))
observer_mat[is.na(observer_mat)] <- "RTM"

sun_mat <- det_covs_matrix %>%
  select(starts_with("time_")) %>%
  as.matrix()
sun_mat[is.na(sun_mat)] <- 0

jul_mat <- det_covs_matrix %>%
  select(starts_with("julian_")) %>%
  as.matrix()
jul_mat[is.na(jul_mat)] <- 0

det_covs_list <- list(
  wind = wind_mat,
  sky = sky_mat,
  observer = observer_mat,
  time_since_dawn = sun_mat,
  julian_day = jul_mat
)

site_covs <- pc %>%
  select(point_id, site_name,year,point) %>%
  distinct() %>%
  arrange(point_id)

site_covs <- site_covs %>%
  filter(point_id %in% site_list) %>%
  arrange(year,factor(point_id, levels = site_list))

site_covs <- site_covs %>%
  select(site_name,year,point)






# check for missing
# Check visits per point_id
visit_counts <- pc %>%
  group_by(point_id, year) %>%
  summarise(n_visits = n_distinct(visit), .groups = "drop")

#print(visit_counts)


### end check


# Abundance and detection formulas
abund_formula <- ~  point+year#site_name # year may have to be a plot-year combo because Nmixture assumes closed populations, so differences between seasons is not accounted for in a meaningful way. maybe
det_formula <- ~ observer+wind + sky  +julian_day#+ time_since_dawn  # this made some species get way overestimated 

# Package data for spAbundance
data_list <- list(
  y = y_array,
  det.covs = det_covs_list,
  abund.covs = site_covs
)

priors <- list(
  # beta is for abundance covariates
  beta.comm.normal = list(0,100),#0, 100), #mean and varaince
  # alpha is for detection covariates
  alpha.comm.normal = list(mean = c(0,    #prior on intercept
                                    0,    # prior on observer
                                    0,
                                    -0.5, #weak negative prior on detection covariats of wind, sky and time since sunrise
                                    -0.5,
                                    -0.5), 
                           var = c(100, #not confident about intercept
                                   100, #not confident about observer
                                   100,
                                   10,  #weakly confident about directionality of wind, sky and time since dawn
                                   10,
                                   10)),
  tau.sq.beta.ig = list(0.1, 0.1), #0.1 is default, value of 1 is more similar, value of 0.01 is pretty uninformative, could vary between species for abundance
  tau.sq.alpha.ig = list(0.1, 0.1),
  sigma.sq.mu.ig = list(0.1, 0.1), #for random effects on abundance
  sigma.sq.p.ig = list(0.1, 0.1)   #for random effects on detection
)

site_covs_expanded <- site_covs[rep(1:n_sites, times = length(unique(pc$year))), ]
n_beta  <- ncol(model.matrix(abund_formula, data = site_covs_expanded))

n_alpha <- 6

inits.list <- list(
  beta.comm = 0,
  alpha.comm = 0,
  beta = matrix(0, nrow = length(species_list), ncol = n_beta),
  alpha = matrix(0, nrow = length(species_list), ncol = n_alpha),
  N = apply(y_array, c(1, 2), max, na.rm = TRUE)#   
)
#dim(apply(y_array, c(1, 2), max, na.rm = TRUE))

# Fit the multi-species N-mixture model
out_nmix <- msNMix(
  abund.formula = abund_formula,
  det.formula = det_formula,
  data = data_list,
  priors = priors,
  tuning = list(
    beta = 0.1,
    alpha = 0.1
  ),
  inits = inits.list,
  n.batch = 1000,  #3000
  batch.length = 25,
  accept.rate = 0.43,
  family = "Poisson",
  n.omp.threads = 1,
  verbose = TRUE,
  n.report = 100,
  n.burn = 500, # 2500
  n.thin = 2,   # 5
  n.chains = 2 # 3
)

globalmod <- (out_nmix)

# Check detection probabilities
summary(out_nmix)  # Look at detection parameter estimates
#det_probs <- predict(out_nmix, type = "detection", X.0 = out_nmix$X, X.p.0 = out_nmix$X.p)
alpha_intercepts <- out_nmix$alpha.samples[, grep("\\(Intercept\\)", colnames(out_nmix$alpha.samples))]

# Mean detection probability by species (at average covariate values)

#mean_det_by_species <- apply(alpha_intercepts, 2, function(x) mean(plogis(x)))
#names(mean_det_by_species) <- out_nmix$sp.names
#print(mean_det_by_species) # NOFL has 0.042 and HAWO has 0.015 for detection prob, removing those


# check trace plot
#plot(out_nmix$beta.samples[, 1], type = "l", ylab = "Parameter value", xlab = "Iteration", main = "Trace plot")



# Extract the summary output

fit_vals <- fitted(out_nmix, type = "marginal")
# Check which species have issues
#fit_vals$y.rep.samples  # Abundance samples by species and site
# Step 1: Sum over visits (4th dim) → iterations × species × site
total_abundance <- apply(fit_vals$y.rep.samples, c(1, 2, 3), sum)

# Step 2: Summarize over iterations for each species × site
n_iter <- dim(total_abundance)[1]
n_species <- dim(total_abundance)[2]
n_sites <- dim(total_abundance)[3]

summary_list <- vector("list", length = n_species * n_sites)
k <- 1
for (s in 1:n_species) {
  for (p in 1:n_sites) {
    values <- total_abundance[, s, p]
    summary_list[[k]] <- data.frame(
      species = species_list[s],
      point_id = site_list[p],
      mean = mean(values),
      lower = quantile(values, 0.025),
      upper = quantile(values, 0.975)
    )
    k <- k + 1
  }
}

df_plot <- do.call(rbind, summary_list)

df_plot$year <- substr(df_plot$point_id, 1, 4)
df_plot <- subset(df_plot,df_plot$year=="2025")
df_plot$point_id <- substring(df_plot$point_id, 5)
df_plot$point_id <- as.numeric(df_plot$point_id)
```

## Summary plots for results thus far...

```{r}
#| echo: false
#| message: false

#plotting


ggplot(df_plot, aes(x = factor(point_id), y = mean, fill = species)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                position = position_dodge(width = 0.9),
                width = 0.3) +
  facet_wrap(~ species) +
  theme_bw() +
  theme(
    axis.title = element_blank(),  # removes x and y axis titles
    legend.position = "none"  # Hide legends from individual plots
  ) +    
  geom_vline(xintercept = 7.5, linetype="dotted", 
             color = "black", linewidth=1.5)+
  geom_vline(xintercept = 14.5, linetype="dotted", 
             color = "black", linewidth=1.5)+
  geom_vline(xintercept = 23.5, linetype="dotted", 
             color = "black", linewidth=1.5)+
  #geom_vline(xintercept = 32.5, linetype="dotted", 
  #           color = "black", linewidth=1.5)+
  #geom_vline(xintercept = 39.5, linetype="dotted", 
  #           color = "black", linewidth=1.5)+
  #geom_vline(xintercept = 46.5, linetype="dotted", 
  #           color = "black", linewidth=1.5)+
  #geom_vline(xintercept = 55.5, linetype="dotted", 
  #           color = "black", linewidth=1.5)+
  ylim(0,30)+
  scale_y_continuous(expand=c(0, 0))+ 
  labs(x = "Point ID", y = "Estimated Abundance", fill = "Species") +
  theme(axis.text.x = element_text(angle = 90))
# to save it call it object g and run the next line
#ggsave("point_count_densities.png",g,dpi=300,height=7,width=10)
```

This plot (above) shows the abundance estimates for cavity nesting species at each study site (separated by vertical dashed lines). This data represents only March and April 2025 since that's what we have entered into excel thus far. Species estimates are shown with their corresponding 95% credible intervals.

# How long should a point count be?

In this study we are doing 10min point counts. It could be that 10min is too long, and most of the species we detect will be detected in a shorter amount of time. If so, we may be able to do shorter point counts but we don't want to risk lowering our detection of birds that are more difficult to detect (those that are quiet or rare). Let's look at a simple species accumulation curve...

```{r}
#| echo: false
#| include: false
#| message: false

#how long is long enough for a point count? an analysis in unmarked


# convert mm:ss to numeric minutes
pc_clean <- pc %>%
  mutate(
    time_mins = as.numeric(ms(time_detected)) / 60  
  )

# Step 1: Keep only the FIRST detection per species per visit (point_id + visit)
pc_unique <- pc_clean %>%
  arrange(point, visit, species, time_mins) %>%
  group_by(point, visit, species) %>%
  slice_head(n = 1) %>%  # keep only first detection per species per visit
  ungroup()

# Step 2: Bin by time again
pc_binned <- pc_unique %>%
  mutate(time_bin = cut(time_mins,
                        breaks = seq(0, 10, by = 2),
                        labels = c("0-2", "2-4", "4-6", "6-8", "8-10"),
                        right = FALSE))

# Step 3: Count number of species per visit per time bin
richness_per_visit <- pc_binned %>%
  group_by(point, visit, time_bin) %>%
  summarise(richness = n_distinct(species), .groups = "drop")

# Step 4: Pivot and calculate cumulative richness
richness_wide <- richness_per_visit %>%
  pivot_wider(names_from = time_bin, values_from = richness, values_fill = 0)


#
# Time bin order
time_bin_order <- c("0-2", "2-4", "4-6", "6-8", "8-10")

# Extract just the time bin values, do row-wise cumsum, and bind back
richness_cum <- richness_wide %>%
  select(point, visit, all_of(time_bin_order)) %>%
  mutate(across(all_of(time_bin_order), as.numeric))  # ensure numeric just in case

# Apply row-wise cumsum to only the bin columns
cumulative_bins <- t(apply(richness_cum[time_bin_order], 1, cumsum))  # returns a matrix

# Bind back to original identifiers
richness_cum_final <- bind_cols(
  richness_wide %>% select(point, visit),
  as.data.frame(cumulative_bins)
)

# Set correct column names
colnames(richness_cum_final)[-(1:2)] <- time_bin_order

# Check
head(richness_cum_final)


#



# Step 5: Average across visits per point
richness_mean_by_plot <- richness_cum_final %>%
  group_by(point) %>%
  summarise(across(`0-2`:`8-10`, mean), .groups = "drop")

# Reshape to long format for plotting
richness_long <- richness_mean_by_plot %>%
  pivot_longer(cols = `0-2`:`8-10`, names_to = "TimeBin", values_to = "MeanRichness")

# Ensure time bins are ordered properly
richness_long$TimeBin <- factor(richness_long$TimeBin, levels = c("0-2", "2-4", "4-6", "6-8", "8-10"))

richness_overall <- richness_long %>%
  group_by(TimeBin) %>%
  summarise(MeanRichness = mean(MeanRichness))



```

```{r}
#| echo: false
#| message: false

#plot the curve
ggplot(richness_long, aes(x = TimeBin, y = MeanRichness, group = factor(point), color = factor(point))) +
  geom_line(alpha = 0.3) +
  geom_line(data = richness_overall, aes(x = TimeBin, y = MeanRichness, group = 1), color = "black", size = 1.5) +
  labs(
    title = "",
    x = "Time Bin (min)",
    y = "Mean Cumulative Species Richness"
  ) +
  scale_y_continuous(breaks = c(0:10),limits = c(0, 10), expand = c(0, 0))+
  theme_bw() +
  theme(legend.position = "none")
```

Looks like modest accumulation with no sharp increase and plateau seen across plots with some variation between plots. The variation between plots could be random noise.

# Point count schedule by month

The following are the scheduled order of point counts to conduct per visit to the site. This ensures that community data obtained isn't biased by lower or higher detections thoughout the day.

## April

### Arcadia

```{r}
#| echo: false
#| message: false
loc <- "ARCA"
mon <- "April"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Arcadia", "in", mon)))
```

### Greenwood

```{r}
#| echo: false
#| message: false
loc <- "GREN"
mon <- "April"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Greenwood", "in", mon)))
```

### Livingston

```{r}
#| echo: false
#| message: false
loc <- "LIVI"
mon <- "April"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Livingston Place", "in", mon)))
```

### Tall Timbers

```{r}
#| echo: false
#| message: false
loc <- "TTRS"
mon <- "April"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Tall Timbers", "in", mon)))
```

## May

### Arcadia

```{r}
#| echo: false
#| message: false
loc <- "ARCA"
mon <- "May"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Arcadia", "in", mon)))
```

### Greenwood

```{r}
#| echo: false
#| message: false
loc <- "GREN"
mon <- "May"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Greenwood", "in", mon)))
```

### Livingston

```{r}
#| echo: false
#| message: false
loc <- "LIVI"
mon <- "May"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Livingston Place", "in", mon)))
```

### Tall Timbers

```{r}
#| echo: false
#| message: false
loc <- "TTRS"
mon <- "May"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Tall Timbers", "in", mon)))
```

## June

### Arcadia

```{r}
#| echo: false
#| message: false
loc <- "ARCA"
mon <- "June"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Arcadia", "in", mon)))
```

### Greenwood

```{r}
#| echo: false
#| message: false
loc <- "GREN"
mon <- "June"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Greenwood", "in", mon)))
```

### Livingston

```{r}
#| echo: false
#| message: false
loc <- "LIVI"
mon <- "June"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Livingston Place", "in", mon)))
```

### Tall Timbers

```{r}
#| echo: false
#| message: false
loc <- "TTRS"
mon <- "June"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Tall Timbers", "in", mon)))
```

## July

### Arcadia

```{r}
#| echo: false
#| message: false
loc <- "ARCA"
mon <- "July"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Arcadia", "in", mon)))
```

### Greenwood

```{r}
#| echo: false
#| message: false
loc <- "GREN"
mon <- "July"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Greenwood", "in", mon)))
```

### Livingston

```{r}
#| echo: false
#| message: false
loc <- "LIVI"
mon <- "July"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Livingston Place", "in", mon)))
```

### Tall Timbers

```{r}
#| echo: false
#| message: false
loc <- "TTRS"
mon <- "July"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Tall Timbers", "in", mon)))
```

## August

### Arcadia

```{r}
#| echo: false
#| message: false
loc <- "ARCA"
mon <- "August"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Arcadia", "in", mon)))
```

### Greenwood

```{r}
#| echo: false
#| message: false
loc <- "GREN"
mon <- "August"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Greenwood", "in", mon)))
```

### Livingston

```{r}
#| echo: false
#| message: false
loc <- "LIVI"
mon <- "August"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Livingston Place", "in", mon)))
```

### Tall Timbers

```{r}
#| echo: false
#| message: false
loc <- "TTRS"
mon <- "August"
month_data <- survey %>%
      filter(location == loc, month_name == mon)
# Print the table for the specific location and month
print(kable(month_data, caption = paste("Order for Tall Timbers", "in", mon)))
```

This page was updated:
```{r}
#| echo: false
#| results: 'asis'

cat(format(Sys.Date(), "%B %d, %Y"))

```